---
title: "Measuring Risk in NYC"
author: Andy Areitio, Claudia Kampbel, Pedro B Franco (INSEAD MBA Class of July 2017);
  Jorge Bravo
date: "January 31, 2017"
output:
  html_document:
    css: Styles/default.css
    theme: paper
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
  pdf_document:
    toc: yes
---

# Project Objective:

We intend to analyze and identify patterns regarding car accidents in New York City ("NYC"). In order to acheive this objective we have obtained from the New York Police Department ("NYPD") open source data with almost two years of accidents. The table has 29 columns containing details on each accident reported to the NYPD in the past 4.5 years.
 
Discalimer: Please note that in order to run this project on a local computer, you will have do download a CSV file from the NYPD web site and name it as follows: 

..Data/NYPD_Motor_Vehicle_Collisions.csv

Link for download:
https://data.cityofnewyork.us/Public-Safety/NYPD-Motor-Vehicle-Collisions/h9gi-nx95

Thank you!

```{r , echo=TRUE, message=FALSE, warning=FALSE}
# basic preparatory commands

# clear the working environment
rm(list = ls())

# load the dataset for the project
NYPDfile <- read.csv("Data/NYPD_Motor_Vehicle_Collisions.csv", header = TRUE, stringsAsFactors = FALSE)


#####

# load required packages
# install required packages if loading failed
# all packages sould be listed inside install_load()
if (!require("ggplot2")) {
  install.packages("ggplot2", repos="http://cran.rstudio.com/") 
  library("ggplot2")
}

###

```


## Project process and document structure

Please find below the process we have used to solve the proposed business problem:

**1. Defining business problem.** 

As described above, our intentions are to analyze and identify patterns concerning car accidents in the city of New York. We intend to draw some conclusions from the data and generate some visualizations of the data.

**2. Find data.** 

Find a good and reliable source of data.

**3. Undestand the data.** 

It is important to understand the data we have at hand before trying to manipulate it. We will describe the columns on the tabels as a part of this process.
Depending on our findings we may have to prepare the data set for modeling:

+ Check for missing values, exclusion of corresponding observations.
+ Check for outliers, decision on their participation in analysis.
+ Conversion of non-numerical attributes to numerical dummy variables.
+ Separation of dataset into training and test.

Part of understanding the data will involve running the following tests:
i.	Analyze data quantitatively (how many zip.codes we have, for example)
ii.	Number of people who die in accidents

Visualizing the data will also be part of this stage, where will try to check if we can see any of the following trends:

i.	Boroughs with most accidents
ii. Boroughs with most deaths
iii.	Boroughs with most injuries
iv. Histograms with the most frequent type of accident (type = # injuries or deaths) 
v. Histograms with accidents per borough

**4. Define business criteria.**  

We will have to define the following:

+ Define a method for rating the accidents (i.e. rows) using URS - Universal Rating System (for accidents)
+ Criteria for dangerous zip codes (scaling URS)

**5. Prepare and model solution.**

+ Choose the columns/criteria that make more sense (dimensionality reduction)
+ Develop logic/workflow
+ Develop code

**6. Test and Deploy.** Test the solution and conclude the project. Conclude the project and add comments on how to improve the solution.

# Business Understanding

Our conclusions can be a base that could be used in the future by insurance companies, people looking for locations to buy houses in NYC, tourists looking for safest routes for walks, or even for waze-type platforms.

# Data Understanding

## Source of data

We have downloaded open source data from the following site:
(https://data.cityofnewyork.us/Public-Safety/NYPD-Motor-Vehicle-Collisions/h9gi-nx95).

## Dataset size and variables

Each line of the table represents an incident involving a motor car vehicle in NYC since July 2012.

Please find below a breif description of the columns:

```{r , echo=TRUE, message=FALSE, warning=FALSE}
str(NYPDfile)

###### JB: maybe to avoid this if the summary is alredy printed below. You could also put only the explanation of variables that
# are not self-explicative 

```

+1. DATE (Floating Time Stamp) - Date of the incident

+2. TIME (Text) - Time the incident occured

+3. BOROUGH (Text) - The Bourough the accident occured

+4. ZIP.CODE (Text) - The Zipcode of the location

+5. LATITUDE (Number #) - Latitude of the location the accident occured

+6. LONGITUDE (Number #) - Longitude of the location the accident occured

+7. LOCATION (Location Datatype) - Location, where logitude and latitude are joined and can be refferenced to a map

+8. ON.STREET.NAME (Plain text) - Name of the street the accident occured                

+9. CROSS.STREET.NAME (Plain text) - Closest cross street           

+10. OFF.STREET.NAME (Plain text)               

+11. NUMBER.OF.PERSONS.INJURED (Number #)    

+12. NUMBER.OF.PERSONS.KILLED (Number #)     

+13. NUMBER.OF.PEDESTRIANS.INJURED (Number #)

+14. NUMBER.OF.PEDESTRIANS.KILLED (Number #)  

+15. NUMBER.OF.CYCLIST.INJURED (Number #)  

+16. NUMBER.OF.CYCLIST.KILLED (Number #)    

+17. NUMBER.OF.MOTORIST.INJURED (Number #)  

+18. NUMBER.OF.MOTORIST.KILLED (Number #)   

+19. CONTRIBUTING.FACTOR.VEHICLE.1 (Plain text) - Main factor which caused the accident (Driver Inexperience, Driver Inattention, Turning Improperly, etc...)

+20. CONTRIBUTING.FACTOR.VEHICLE.2 (Plain text)

+21. CONTRIBUTING.FACTOR.VEHICLE.3 (Plain text)

+22. CONTRIBUTING.FACTOR.VEHICLE.4 (Plain text)

+23. CONTRIBUTING.FACTOR.VEHICLE.5 (Plain text)

+24. UNIQUE.KEY  (Number #) 

+25. VEHICLE.TYPE.CODE.1 (Plain text) - Type of Vehicle 1 (sports utility, taxi, passenger, bus, etc)         

+26. VEHICLE.TYPE.CODE.2 (Plain text)          

+27. VEHICLE.TYPE.CODE.3 (Plain text)         

+28. VEHICLE.TYPE.CODE.4 (Plain text)          

+29. VEHICLE.TYPE.CODE.5 (Plain text)    

First, let's generate summary report for all variables.

+Below is summary of project dataset structure.

```{r , echo=TRUE, message=FALSE, warning=FALSE}

#colnames(NYPDfile)
nrow(NYPDfile)
ncol(NYPDfile)
summary(NYPDfile)
```

What can you see in the summary is that there are a lot of NA's in certain data which could be difficult to use. We will visualize the data to find out more.

'Date' and 'Time' columns are defined as character and not numerical. These variables are not key to our primary analysis (assessing the risk of having an accident in a specified ZIP code/borough) so we will focus on the remaining variables.


```{r , echo=TRUE, message=FALSE, warning=FALSE}

#analyzing categorical variables and unique values
sapply(NYPDfile, function(x) length(unique(x)))

#zooming in one column
 unique(NYPDfile$BOROUGH)
 
 ### JB: we see that the main variable to use is the Borough. However Borough has a lot of unknowns.  zip code is complicated due to the big number (maybe a way of splitting if the zip codes in NY follow a rule and they can be used to disaggreagte the information of a Borough). Cross.Street.Name and Off.Stree.name have too many categories and Contributing.facto.Vehicles also in that case. Vehicle.Type.Code why not?

```


```{r , echo=TRUE, message=FALSE, warning=FALSE}


NYPDfile$BOROUGH[NYPDfile$BOROUGH == ""] <- "UNSPECIFIED"
NYPDfile$ZIP.CODE[is.na(NYPDfile$ZIP.CODE)] <- "00000"


#histogram

NYPDdata <- NYPDfile$UNIQUE.KEY
NYPDdata <- cbind(NYPDdata, NYPDfile$BOROUGH)
colnames(NYPDdata) <- c("ID", "BOROUGH")

NYPDdata[,"BOROUGH"] <- replace(NYPDdata[,"BOROUGH"], NYPDdata[,"BOROUGH"]=="UNSPECIFIED", 0)
NYPDdata[,"BOROUGH"] <- replace(NYPDdata[,"BOROUGH"], NYPDdata[,"BOROUGH"]=="BRONX", 1)
NYPDdata[,"BOROUGH"] <- replace(NYPDdata[,"BOROUGH"], NYPDdata[,"BOROUGH"]=="BROOKLYN", 2)
NYPDdata[,"BOROUGH"] <- replace(NYPDdata[,"BOROUGH"], NYPDdata[,"BOROUGH"]=="MANHATTAN", 3)
NYPDdata[,"BOROUGH"] <- replace(NYPDdata[,"BOROUGH"], NYPDdata[,"BOROUGH"]=="QUEENS", 4)
NYPDdata[,"BOROUGH"] <- replace(NYPDdata[,"BOROUGH"], NYPDdata[,"BOROUGH"]=="STATEN ISLAND", 5)

NYPDdata<-matrix(as.numeric(unlist(NYPDdata)),nrow=nrow(NYPDdata))
colnames(NYPDdata) <- c("ID", "BOROUGH")

hist(NYPDdata[,"BOROUGH"])
```

With

0 = UNSPECIFIED
1 = BRONX
2 = BROOKLYN
3 = MANHATTAN
4 = QUEENS
5 = STATEN ISLAND

The histogram shows that there are a lot of emtpy cells in the BOROUGH column. 

When solving this question, we realize that when the Borough value is an unknown, the zip code is also unknown. When this is the case, the only values regarding location that are sometimes available are On.Street.Name, Cross.Street.Name and Off.Street.Name. 

To assess how many rows are complete the next step is to count how many empty values we have in the columns Borough.

```{r echo=FALSE}

amountUNSPECIFIED <- length(NYPDfile$BOROUGH[NYPDfile$"BOROUGH"=="UNSPECIFIED"])
amountUNSPECIFIED

```

For the column ZIP code, we have the following amount of empty cells:
```{r echo=FALSE}

amount00000 <- length(NYPDfile$ZIP.CODE[NYPDfile$"ZIP.CODE"=="00000"])
amount00000

```
The difference between the 2 values is not significant. We are assuming that always both of the cells are empty.  In this case, the difference between the # of empty ZIP.CODEs and BOROUGHs is only 5 values, we assume that we have a ZIP.CODE value for every row that has a BOROUGH value.

In total, there is a very low percentage of complete rows:

```{r echo=FALSE}

amountUNSPECIFIED/(nrow(NYPDfile))*100

```
Percent of the data does not include values for the column ZIP.CODE AND BOROUGH.  

In absolute terms we have a very large sample that we can analyze, containing over 712 652 accidents  (this tranlates into ~13200 accidents per month). For this analysis we will focus on our sample of completed rows including BOROUGH and ZIP.CODE data. Therefore,we will exclude all rows with missing ZIP.CODES or BOROUGH values. 

## Dataset visualization

1. We will now plot a graph with accidents per borough (independent of time and date):

```{r}


NYPDfile$BOROUGH[NYPDfile$BOROUGH == ""] <- "UNSPECIFIED"
NYPDfile$ZIP.CODE[is.na(NYPDfile$ZIP.CODE)] <- "00000"


#histogram

NYPDdata <- NYPDfile$UNIQUE.KEY
NYPDdata <- cbind(NYPDdata, NYPDfile$BOROUGH)
colnames(NYPDdata) <- c("ID", "BOROUGH")

#NYPDdata[,"BOROUGH"] <- replace(NYPDdata[,"BOROUGH"], NYPDdata[,"BOROUGH"]=="UNSPECIFIED", 0)
NYPDdata[,"BOROUGH"] <- replace(NYPDdata[,"BOROUGH"], NYPDdata[,"BOROUGH"]=="BRONX", 1)
NYPDdata[,"BOROUGH"] <- replace(NYPDdata[,"BOROUGH"], NYPDdata[,"BOROUGH"]=="BROOKLYN", 2)
NYPDdata[,"BOROUGH"] <- replace(NYPDdata[,"BOROUGH"], NYPDdata[,"BOROUGH"]=="MANHATTAN", 3)
NYPDdata[,"BOROUGH"] <- replace(NYPDdata[,"BOROUGH"], NYPDdata[,"BOROUGH"]=="QUEENS", 4)
NYPDdata[,"BOROUGH"] <- replace(NYPDdata[,"BOROUGH"], NYPDdata[,"BOROUGH"]=="STATEN ISLAND", 5)

NYPDdata<-matrix(as.numeric(unlist(NYPDdata)),nrow=nrow(NYPDdata))
colnames(NYPDdata) <- c("ID", "BOROUGH")

hist(NYPDdata[,"BOROUGH"])

```


With

1 = BRONX
2 = BROOKLYN
3 = MANHATTAN
4 = QUEENS
5 = STATEN ISLAND

*According to our sample, the boroughs with the highest number of accidents are Brooklyn and Queens.* 

2. Generate tables with accidents per borough / per type of victim (cyclist / Pedestrian / Motorist):

a. Total number of people killed by BOROUGH (Cyclists+Pededstrians+Motorists)
b. Total number of people injured by BOROUGH (Cyclists+Pededstrians+Motorists)
c. Number of cyclists killed by BOROUGH
d. Number of cyclists injured by BOROUGH
e. Number of pedestrians killed by BOROUGH
f. Number of pedestrians injured by BOROUGH
g. Number of Motorists killed by BOROUGH
h. Number of Motorists injured by BOROUGH

```{r , echo=TRUE, message=FALSE, warning=FALSE}

#histogram


PERSONS.K <- aggregate(NYPDfile$NUMBER.OF.PERSONS.KILLED, list(Borough = NYPDfile$BOROUGH), sum)
PERSONS.I <- aggregate(NYPDfile$NUMBER.OF.PERSONS.INJURED, list(Borough = NYPDfile$BOROUGH), sum)
CYCLISTS.K <- aggregate(NYPDfile$NUMBER.OF.CYCLIST.KILLED, list(Borough = NYPDfile$BOROUGH), sum)
CYCLISTS.I <- aggregate(NYPDfile$NUMBER.OF.CYCLIST.INJURED, list(Borough = NYPDfile$BOROUGH), sum)
PED.K <- aggregate(NYPDfile$NUMBER.OF.PEDESTRIANS.KILLED, list(Borough = NYPDfile$BOROUGH), sum)
PED.I <- aggregate(NYPDfile$NUMBER.OF.PEDESTRIANS.INJURED, list(Borough = NYPDfile$BOROUGH), sum)
MOT.K <- aggregate(NYPDfile$NUMBER.OF.MOTORIST.KILLED, list(Borough = NYPDfile$BOROUGH), sum)
MOT.I <- aggregate(NYPDfile$NUMBER.OF.MOTORIST.INJURED, list(Borough = NYPDfile$BOROUGH), sum)


PERSONS.K
PERSONS.I
CYCLISTS.K
CYCLISTS.I
PED.K
PED.I
MOT.K
MOT.I

##NYPDdata<-matrix(as.numeric(unlist(NYPDdata)),nrow=nrow(NYPDdata))
##colnames(NYPDdata) <- c("ID", "BOROUGH")

##hist(NYPDdata[,"BOROUGH"])
```

From this data we can draw the following conclusions:
a. Brooklyn and Queens have the largest amount of deaths and injures (in total)
b. However, there are more cyclists injured in Manhattan than in Queens (maybe people cycle more in Manhattan than in Queens)
c. Similarly, there are more pedestrians injured in Manhattan than in queens (maybe people like walking a lot more in Manhattan - makes sense!)
d. For motorists, Brooklyn, Queens and Bronx have the largest amount of deaths and injuries (Makes sense as in Manhattan cars can't go very fast) 

3. Generate a histogram with the type of vehicle which was the responsible for the accident:

```{r , echo=TRUE, message=FALSE, warning=FALSE}


NYPDfile$VEHICLE.TYPE.CODE.1[NYPDfile$VEHICLE.TYPE.CODE.1 == ""] <- "0000"

#histogram

NYPDdata1 <- NYPDfile$UNIQUE.KEY
NYPDdata1 <- cbind(NYPDdata1, NYPDfile$VEHICLE.TYPE.CODE.1)
colnames(NYPDdata1) <- c("ID", "VEHICLE")

#NYPDdata1[,"VEHICLE"] <- replace(NYPDdata1[,"VEHICLE"], NYPDdata1[,"VEHICLE"]=="0000", 0)
NYPDdata1[,"VEHICLE"] <- replace(NYPDdata1[,"VEHICLE"], NYPDdata1[,"VEHICLE"]=="PASSENGER VEHICLE", 1)
NYPDdata1[,"VEHICLE"] <- replace(NYPDdata1[,"VEHICLE"], NYPDdata1[,"VEHICLE"]=="SPORT UTILITY / STATION WAGON", 2)
NYPDdata1[,"VEHICLE"] <- replace(NYPDdata1[,"VEHICLE"], NYPDdata1[,"VEHICLE"]=="TAXI", 3)
NYPDdata1[,"VEHICLE"] <- replace(NYPDdata1[,"VEHICLE"], NYPDdata1[,"VEHICLE"]=="MOTORCYCLE", 4)
NYPDdata1[,"VEHICLE"] <- replace(NYPDdata1[,"VEHICLE"], NYPDdata1[,"VEHICLE"]=="PICK-UP TRUCK", 5)
NYPDdata1[,"VEHICLE"] <- replace(NYPDdata1[,"VEHICLE"], NYPDdata1[,"VEHICLE"]=="OTHER", 6)
NYPDdata1[,"VEHICLE"] <- replace(NYPDdata1[,"VEHICLE"], NYPDdata1[,"VEHICLE"]=="VAN", 7)
NYPDdata1[,"VEHICLE"] <- replace(NYPDdata1[,"VEHICLE"], NYPDdata1[,"VEHICLE"]=="BICYCLE", 8)
NYPDdata1[,"VEHICLE"] <- replace(NYPDdata1[,"VEHICLE"], NYPDdata1[,"VEHICLE"]=="SMALL COM VEH(4 TIRES)", 9)
NYPDdata1[,"VEHICLE"] <- replace(NYPDdata1[,"VEHICLE"], NYPDdata1[,"VEHICLE"]=="AMBULANCE", 10)
NYPDdata1[,"VEHICLE"] <- replace(NYPDdata1[,"VEHICLE"], NYPDdata1[,"VEHICLE"]=="LARGE COM VEH(6 OR MORE TIRES)", 11)
NYPDdata1[,"VEHICLE"] <- replace(NYPDdata1[,"VEHICLE"], NYPDdata1[,"VEHICLE"]=="UNKNOWN", 12)
NYPDdata1[,"VEHICLE"] <- replace(NYPDdata1[,"VEHICLE"], NYPDdata1[,"VEHICLE"]=="LIVERY VEHICLE", 13)
NYPDdata1[,"VEHICLE"] <- replace(NYPDdata1[,"VEHICLE"], NYPDdata1[,"VEHICLE"]=="BUS", 14)
NYPDdata1[,"VEHICLE"] <- replace(NYPDdata1[,"VEHICLE"], NYPDdata1[,"VEHICLE"]=="SCOOTER", 15)
NYPDdata1[,"VEHICLE"] <- replace(NYPDdata1[,"VEHICLE"], NYPDdata1[,"VEHICLE"]=="FIRE TRUCK", 16)
NYPDdata1[,"VEHICLE"] <- replace(NYPDdata1[,"VEHICLE"], NYPDdata1[,"VEHICLE"]=="PEDICAB", 17)

NYPDdata1<-matrix(as.numeric(unlist(NYPDdata1)),nrow=nrow(NYPDdata1))
colnames(NYPDdata1) <- c("ID", "VEHICLE")

hist(NYPDdata1[,"VEHICLE"])
```

Where:
1 PASSENGER VEHICLE
2 SPORT UTILITY / STATION WAGON
3 TAXI
4 MOTORCYCLE
5 PICK-UP TRUCK
6 OTHER
7 VAN
8 BICYCLE
9 SMALL COM VEH(4 TIRES)
10 AMBULANCE
11 LARGE COM VEH(6 OR MORE TIRES)
12 UNKNOWN
13 LIVERY VEHICLE
14 BUS
15 SCOOTER
16 FIRE TRUCK
17 PEDICAB 

From our new histogram, we can conclude that passanger vehicles and sport utilities are the ones involved with the highest # of accidents.

4. Next, we will plot simple graphs (independent of time and date and ordered from largest # of accidents to smallest): 

a. Accidents per zip code
b. Deaths per zip code
c. Injuries per zip code

```{r}

PERSONS.K.Z <- aggregate(NYPDfile$NUMBER.OF.PERSONS.KILLED, list(ZIP.CODE = NYPDfile$ZIP.CODE), sum)
PERSONS.I.Z <- aggregate(NYPDfile$NUMBER.OF.PERSONS.INJURED, list(ZIP.CODE = NYPDfile$ZIP.CODE), sum)
CYCLISTS.K.Z <- aggregate(NYPDfile$NUMBER.OF.CYCLIST.KILLED, list(ZIP.CODE = NYPDfile$ZIP.CODE), sum)
CYCLISTS.I.Z <- aggregate(NYPDfile$NUMBER.OF.CYCLIST.INJURED, list(ZIP.CODE = NYPDfile$ZIP.CODE), sum)
PED.K.Z <- aggregate(NYPDfile$NUMBER.OF.PEDESTRIANS.KILLED, list(ZIP.CODE = NYPDfile$ZIP.CODE), sum)
PED.I.Z <- aggregate(NYPDfile$NUMBER.OF.PEDESTRIANS.INJURED, list(ZIP.CODE = NYPDfile$ZIP.CODE), sum)

PERSONS.K.Z
PERSONS.I.Z
CYCLISTS.K.Z
CYCLISTS.I.Z
PED.K.Z
PED.I.Z

```

From the data we can see that 11203 ZIP.CODE (Canarsie) has the highest number of accidents which result in death. Its positioned in the outskirts of NYC in Brooklyn.

## Define Business Criteria

In an attempt to create a "traffic light" system for the zip codes in New York we decided to rate every accident using the following criteria:
+ Every row is a collision and receives 1 point
+ Every injury the row receives 10 points (# of injuries in the row * 3) - Column name do be used: 11. NUMBER.OF.PERSONS.INJURED    
+ Every death the row receives 1000 points - Column name to be used: 12. NUMBER.OF.PERSONS.KILLED 

A new column with the above point system will be created and called ("URS" = Universal Rating System (for accidents))

```{r}

NYPDfile$URS <- NYPDfile$NUMBER.OF.PERSONS.KILLED*2000+NYPDfile$NUMBER.OF.PERSONS.INJURED*10+1

```

A. Here is a simple table of URS per borough

```{r}

aggregate(NYPDfile$URS, list(Borough = NYPDfile$BOROUGH), sum)


```

From this table, we can see that in our sample, Brooklyn is the most dangerous borough according to the URS scale. 

We can also assess the average URS of an accident per borough: 

```{r}

aggregate(NYPDfile$URS, list(Borough = NYPDfile$BOROUGH), mean)


```
From this table we can see that in our sample, on average, the accidents that happen in the Bronx are the most dangerous ones (URS=6.33). In comparison, the accidents that happen in Manhattan are much less dangerous (only URS=4.32)

This is interesting: The Unspecified accidents have the highest URS index, therefore we can infer, as a hypothesis, that many of the most dangerous accidents are not properly classified. Maybe because the police men are worried with the outcome, emergencies, etc... 

Brooklyn is the most dangerous borough and Bronx has the second highest URS, although it was forth in the number of accidents. This suggests that people drive faster or more agressively in the Bronx.



***